Generating Random Logic Programs Using Constraint Programming

### 2020-01-24 paper/paper.tex
(Most of the) conclusion

% \begin{figure}[h]
% \vspace{1in}
% \caption{Sample Figure Caption}
% \end{figure}

% \begin{table}[h]
% \caption{Sample Table Title}
% \label{sample-table}
% \begin{center}
% \begin{tabular}{ll}
% \multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION} \\
% \hline \\
% Dendrite         &Input terminal \\
% Axon             &Output terminal \\
% Soma             &Cell body (contains cell nucleus) \\
% \end{tabular}
% \end{center}
% \end{table}

% UAI format: section/subsection titles are in all caps, figure/table titles are
% in TitleCase, table caption above the table

@@

%TODO: define all the relevant terminology from logic and constraint
% programming: predicates ,arities, (logic) variables, constants, clauses, atoms,
% heads, bodies,
% CP: decision variables, constraints.
% TODO: need to introduce the idea of predicate independence

@@

We also define a measure of how complicated a body of a clause can become. As
each body is represented by a tree (see \cref{sec:bodies}), we set
$\maxNumNodes{} \ge 1$ to be the maximum number of nodes in the tree
representation of any clause. We also set $\maxNumClauses{}$ to be the maximum
number of clauses in a program. We must have that $\maxNumClauses{} \ge
|\predicates{}|$ because we require each predicate to have at least one clause
that defines it. The model supports eliminating either all cycles or just
negative cycles (see \cref{sec:cycles}) and enforcing predicate independence
(see \cref{sec:independence}), so a list of independent pairs of predicates is
another parameter. Since this model can generate probabilistic as well as
non-probabilistic programs, each clause is paired with a probability which is
randomly selected from a given list (which is our last parameter). For
generating non-probabilistic programs, one can set this list equal to $\{ 1 \}$.
Finally, we define $\tokens{} = \{ \neg, \land, \lor, \top \}$ as the set of
tokens that (together with atoms) form a clause. All decision variables of the
model can now be divided into $2 \times \maxNumClauses{}$ separate groups,
treating the body and the head of each clause separately. We say that the
variables are contained in two arrays: $\arrayt{bodies}{\maxNumClauses{}}{Body}$
and $\arrayt{heads}{\maxNumClauses{}}{Head}$. Since the order of the clauses
does not change the meaning of the program, we can also state our first
constraint:
\begin{constraint}
  Clauses are sorted.
\end{constraint}
Here and henceforth, the exact ordering is immaterial: we only impose an order
to eliminate permutation symmetries.

@@

\begin{definition}
  The \emph{head} of a clause is composed of a $\variable{predicate} \in
  \predicates \cup \{ \Box \}$, and
  $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}} \cup \{ \Box
  \}$.
\end{definition}
Here, we use $\Box$ to denote either a disabled clause that we choose not to use
or disabled arguments if the arity of the $\variable{predicate}$ is less than
$\maxArity{}$. The reason why we need a separate value for the latter (i.e., why
it is not enough to fix disabled arguments to a single already-existing value)
will become clear in \cref{sec:variable_symmetry}.

@@

\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let
  \[
    S = \{ \maxArity{} \times (i + 1) + j + 1 \mid j = 0, \dots, \maxArity{} - 1.
    \}
  \]
  If $\variable{values}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntros} \cap S = \emptyset.
  \]
\end{constraint}
In simpler terms, \cref{constraint:potentialIntroductions} says that if a node
in the tree representation of a clause represents something other than a
predicate, then a variable cannot be introduced as one of its `arguments'.

@@

We will first consider clauses with gaps, i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and total arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\begin{align*}
  T(n, a) = T(n-1, a) + 2&\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
  c_i \ge 1 \text{ for all } i}}\\
  &\sum_{\substack{d_1 + \dots + d_k = a,\\
  d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\end{align*}
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

We can now add a constraint to define an adjacency matrix for the predicate
dependency graph but without positivity/negativity.
\begin{definition} \label{def:adjacency_matrix}
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by stating that $\mathbf{A}[i][j] = 0$
  if and only if, for all $k \in \{ 0, \dots, \maxNumClauses{} - 1 \}$, either
  \[
    \variable{heads}[k].\variable{predicate} \ne j
  \]
  or
  \[
    i \not\in \{a.\variable{name} \mid a \in
    \variable{bodies}[k].\variable{values} \}.
  \]
\end{definition}

@@

\begin{figure}
  \centering
  \[
    \begin{blockarray}{ccccc}
      \begin{block}{c(cccc)}
        \mathsf{father} & 0 & 0 & 0 & 0 \\
        \mathsf{mother} & 1 & 0 & 0 & 0 \\
        \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
        \mathsf{sibling} & 0 & 0 & 0 & 0 \\
      \end{block}
    \end{blockarray}
  \]
  \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for
    \cref{example:independence}}
  \label{fig:dependencies_matrix}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth}] (parent) edge (father);
    \draw[-{Stealth}] (mother) edge (father);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
  \end{tikzpicture}
  \caption{The predicate dependency graph that corresponds to
    \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
    or may not exist.}
  \label{fig:dependencies2}
\end{figure}

@@

% TODO: talk about randomness. The value heuristic is random because we want
% randomness. After finding one solution, one can continue the search for more,
% but they will be similar (acceptable if the search space is small and we want
% to find all solutions).

\section{CONCLUSION \& FUTURE WORK}

\begin{itemize}
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
%\item Inference options to explore. Logspace vs normal space. Symbolic vs
%  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
%  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
%  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible.
\item Could add statistics about what constraints tend to conflict.
\end{itemize}

>>>>

@@

%TODO: define all the relevant terminology from logic and constraint
% programming: predicates ,arities, (logic) variables, constants, clauses, atoms,
% heads, bodies, formulas.
% CP: decision variables, constraints. So the word 'variable' has two meanings.
% TODO: need to introduce the idea of predicate independence

@@

We also define a measure of how complicated a body of a clause can become. As
each body is represented by a tree (see \cref{sec:bodies}), we set
$\maxNumNodes{} \ge 1$ to be the maximum number of nodes in the tree
representation of any clause. We also set $\maxNumClauses{}$ to be the maximum
number of clauses in a program. We must have that $\maxNumClauses{} \ge
|\predicates{}|$ because we require each predicate to have at least one clause
that defines it. The model supports eliminating either all cycles or just
negative cycles (see \cref{sec:cycles}) and enforcing predicate independence
(see \cref{sec:independence}), so a list of independent pairs of predicates is
another parameter. Since this model can generate probabilistic as well as
non-probabilistic programs, each clause is paired with a probability which is
randomly selected from a given list (which is our last parameter). For
generating non-probabilistic programs, one can set this list equal to $\{ 1 \}$.
Finally, we define $\tokens{} = \{ \neg, \land, \lor, \top \}$ as the set of
tokens that (together with atoms) form a clause. All decision variables of the
model can now be divided into $2 \times \maxNumClauses{}$ separate groups,
treating the body and the head of each clause separately. We say that the
variables are contained in two arrays: $\arrayt{bodies}{\maxNumClauses{}}{Body}$
and $\arrayt{heads}{\maxNumClauses{}}{Head}$. Since the order of the clauses
does not change the meaning of the program, we can also state our first
constraint:

\begin{constraint}
  Clauses are sorted.
\end{constraint}

Here and henceforth, the exact ordering is immaterial: we only impose an order
to eliminate permutation symmetries.

@@

\begin{definition}
  The \emph{head} of a clause is composed of a $\variable{predicate} \in
  \predicates \cup \{ \Box \}$, and
  $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}} \cup \{ \Box
  \}$.
\end{definition}

Here, we use $\Box$ to denote either a disabled clause that we choose not to use
or disabled arguments if the arity of the $\variable{predicate}$ is less than
$\maxArity{}$. The reason why we need a separate value for the latter (i.e., why
it is not enough to fix disabled arguments to a single already-existing value)
will become clear in \cref{sec:variable_symmetry}.

@@

\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let
  \[
    S = \{ \maxArity{} \times (i + 1) + j + 1 \mid j = 0, \dots, \maxArity{} - 1.
    \}
  \]
  If $\variable{values}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntros} \cap S = \emptyset.
  \]
\end{constraint}

In simpler terms, \cref{constraint:potentialIntroductions} says that if a node
in the tree representation of a clause represents something other than a
predicate, then a variable cannot be introduced as one of its `arguments'.

@@

We will first consider clauses with gaps, i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and total arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\begin{align*}
  T(n, a) = T(n-1, a) + 2&\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
  c_i \ge 1 \text{ for all } i}}\\
  &\sum_{\substack{d_1 + \dots + d_k = a,\\
  d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\end{align*}
The first term here represents negation, i.e., negating a formula consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

We can now add a constraint to define an adjacency matrix for the predicate
dependency graph but without positivity/negativity.

\begin{definition} \label{def:adjacency_matrix}
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by stating that $\mathbf{A}[i][j] = 0$
  if and only if, for all $k \in \{ 0, \dots, \maxNumClauses{} - 1 \}$, either
  \[
    \variable{heads}[k].\variable{predicate} \ne j
  \]
  or
  \[
    i \not\in \{a.\variable{name} \mid a \in
    \variable{bodies}[k].\variable{values} \}.
  \]
\end{definition}

@@

\begin{figure}
  \centering
  \[
    \begin{blockarray}{ccccc}
      \begin{block}{c(cccc)}
        \mathsf{father} & 0 & 0 & 0 & 0 \\
        \mathsf{mother} & 1 & 0 & 0 & 0 \\
        \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
        \mathsf{sibling} & 0 & 0 & 0 & 0 \\
      \end{block}
    \end{blockarray}
  \]
  \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for
    \cref{example:independence}}
  \label{fig:dependencies_matrix}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth}] (parent) edge (father);
    \draw[-{Stealth}] (mother) edge (father);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
    \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
  \end{tikzpicture}
  \caption{The predicate dependency graph that corresponds to
    \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
    or may not exist.}
  \label{fig:dependencies2}
\end{figure}

@@

% TODO: talk about randomness. The value heuristic is random because we want
% randomness. After finding one solution, one can continue the search for more,
% but they will be similar (acceptable if the search space is small and we want
% to find all solutions).
% Mention the random heuristic. Mention that restarting gives better randomness,
% but duplicates become possible.
% TODO: the number of predicates is also important.

\section{CONCLUSIONS}

% TODO: the positive side (what did we achieve?)

One major question that was not addressed in this paper is that of logical
equivalence. We eliminated permutation symmetries and constrained variable usage
so that the same `functionality' is always assigned to the same variable, but
syntactically different logical formulas are always treated as being different.
For example, all three of these formulas are logically equivalent but
syntactically different: $\neg(\mathsf{P}(X) \lor \mathsf{Q}(X))$,
$\neg\mathsf{P}(X) \land \neg\mathsf{Q}(X)$, and $\neg\mathsf{Q}(X) \land
\neg\mathsf{P}(X)$. While there are many situations where one would prefer to
treat logically equivalent formulas as symmetries that should be eliminated,
there are also situations where we do care about enumerating different ways to
express the same probability distribution or knowledge base. For instance,
one could look for the `best' logically equivalent formula, or investigate
whether some formulas result in faster inference than others.

Another issue worth discussing is that of randomness. Most random models sample
from a well-defined probability distribution whereas a constraint model, while
potentially sufficiently random, provides no such guarantees. In future work, it
would be interesting to examine the probability distribution of logic programs
as generated by our model, although it is not obvious how such a
probability distribution could be characterised. One example of a clear bias
exhibited by our model is that of complexity over simplicity, i.e., if a tree
that represents the body of a clause can have up to $\maxNumNodes{}$
nodes---because the number of possible trees with $n$ nodes grows with $n$---it
is much more likely to have $\maxNumNodes{}$ or $\maxNumNodes{} - 1$ nodes than,
e.g., one or two.
